---
output:
  pdf_document: default
  html_document: default
---
Loading and Preprocessing data
```{r}
data <- read.csv("activity.csv")
data$date<-as.Date(data$date,format = "%Y-%m-%d")

```

Total number of steps taken per day
```{r message=FALSE, warning=FALSE}
library(dplyr)
stepsperday <- data %>% 
        group_by(date) %>%
        summarise(steps_per_day=sum(steps, na.rm = T))

```

Histogram of the total number of steps taken each day


```{r}
hist(stepsperday$steps_per_day,
     xlab ="Steps per day",
     main = "Histogram of total number of steps taken each day"
)

```

Calculation of mean and median of the total number of steps taken per day

```{r}
mn <- mean(stepsperday$steps_per_day)
mdn <- median(stepsperday$steps_per_day)

```

The mean of total steps taken per day is `r mn` and the median `r mdn`



```{r}
AvrgStepsPerInt <- data %>%
        group_by(interval) %>%
        summarise(AverageSteps=mean(steps,na.rm = T))

plot(AvrgStepsPerInt$interval,
     AvrgStepsPerInt$AverageSteps, 
     type="l", 
     xlab="Interval",
     ylab= "Average Steps", 
     main="Average Steps per 5 min interval"
)


```

Calculation of 5-minute interval which has on average across all the days,the maximum number of steps


```{r}
maxstepsinterval <- which(AvrgStepsPerInt$AverageSteps==max(AvrgStepsPerInt$AverageSteps))


```

The 5-minute interval which has on average the maximum number of steps is `r AvrgStepsPerInt$interval[maxstepsinterval]`


Calculation of the total number of missing values in the dataset

```{r}

missingValues <- sum(is.na(data))

```

The total number of missing values is `r missingValues`

Filling of missing Values


```{r}
#creating new dataset to be filled 
FilledDataset <- data

for (i in 1:nrow(FilledDataset)){
        
        #finding the average steps per 5 minute interval of the corresponding day
        IntIndic <- which(AvrgStepsPerInt$interval==FilledDataset$interval[i])
        
        if (is.na(FilledDataset$steps[i] == T)) {
                FilledDataset$steps[i] <- AvrgStepsPerInt$AverageSteps[IntIndic]
        }
}

```



```{r}

stepsperdayfilled <- FilledDataset %>% 
        group_by(date) %>%
        summarise(steps_per_day=sum(steps, na.rm = T))


hist(stepsperdayfilled$steps_per_day,
     xlab ="Steps per day",
     main = "Histogram of total number of steps taken each day (NAs Filled)"
)

```

Calculation the mean and median of the new dataset
```{r}
mnfld <- mean(stepsperdayfilled$steps_per_day)
mdnfld <- median(stepsperdayfilled$steps_per_day)

```


The mean and median of total number of steps taken each day, after the NAs were filled is `r sprintf("%.2f",mnfld) `, `r sprintf("%.2f",mdnfld)` respectively. Both values changed from the estimates of the first part. The imputing of missing data moved the median a little higher but had bigger effect on the mean value. It is worthy to note that they have the same value after the imputation.



Creating a new variable which indicates if the corresponding day is weekend or weekday

```{r}
FilledDataset$TypeOfDay <- "Weekday"

FilledDataset$TypeOfDay[weekdays(FilledDataset$date) %in% c("Saturday" , "Sunday")] <- "Weekend"

FilledDataset$TypeOfDay <- as.factor(FilledDataset$TypeOfDay)


```


Creating the two plots to compare the activity of weekdays vs weekends
```{r message=FALSE, warning=FALSE}
AvrgStepsPerIntFilled <- FilledDataset %>%
        group_by(TypeOfDay, interval) %>%
        summarise(AverageSteps=mean(steps)) %>% 
        ungroup()


wkdayIndic <- AvrgStepsPerIntFilled$TypeOfDay=="Weekday"
wkendIndic <- AvrgStepsPerIntFilled$TypeOfDay=="Weekend"

par(mfrow=c(2,1),ann=F)

plot(AvrgStepsPerIntFilled$interval[wkdayIndic],
     AvrgStepsPerIntFilled$AverageSteps[wkdayIndic],
     type = "l",
     )
mtext("Weekdays", side=4, cex=1)
mtext("Average Steps - Weekdays vs Weekend", side=3, cex=1.5, line=1)

plot(AvrgStepsPerIntFilled$interval[wkendIndic],
     AvrgStepsPerIntFilled$AverageSteps[wkendIndic],
     type = "l"
     )

mtext("Weekend", side=4, cex=1)
mtext("Interval", side=1, cex=1.2,line= 2.5)

```



From the two plots we can observe that people tend to start their day earlier during weekdays. Both of them are fluctuating but during weekends there is a significant higher level of activity throughout the day 





